---
title: "The Great Coffee Caper"
subtitle: "Exploring Discernment and Appreciation of Different Roasts"
author: 
  - name: "Cristina Lafuente"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Evaluation of data recovered from coffee taste test"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  message: FALSE
  warning: FALSE
---

```{r}
#| label: load-pkgs
#| message: FALSE


if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(cowplot,
               dplyr,
               here,
               knitr,
               magick,
               scales,
               tidyverse,
               viridis)

devtools::install_github("tidyverse/dsbox")
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

coffeeData <- readr::read_csv(('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-05-14/coffee_survey.csv'))
```

The Coffee Data comes from a coffee taste test held on YouTube in Oct 2023. Data was self reported by participants. It was aggregated by a data blogger named Robert McKeon Aloe who analyzed the data the following month.

The Dimensions of the data are:

```{r}
dim(coffeeData)
```

The variables collected are:

```{r}
colnames(coffeeData)
```

The reason I selected this data set is because I enjoy coffee very much! Too much? Probably not. I have 6 different coffee brewing apparatuses in my kitchen.

```{r fig.cap = "Images from amazon.com reproduced here without permission"}

mokaPot <- (here("images/mokaPot.png"))
espressoMaker <- (here("images/espressoMachine.png"))
frenchPress <- (here("images/frenchPress.png"))
pod <- (here("images/kCup.png"))
percolator <- (here("images/percolator.png"))
pourOver <- (here("images/pourOver.png"))

ggdraw() +
  draw_image(mokaPot, width = 0.33, height = .5, y = 0, interpolate = TRUE) +
  draw_image(espressoMaker, width = 0.33, height = 0.5, x = 0.33, y = 0, interpolate = TRUE) +
  draw_image(frenchPress, width = 0.33, height = 0.5, x = 0.67, y = 0, interpolate = TRUE) + 
  draw_image(pod, width = 0.33, height = 0.5, y = 0.5 , interpolate = TRUE) +
  draw_image(percolator, width = 0.33, height = 0.5, x = 0.33, y = 0.5, interpolate = TRUE) + 
  draw_image(pourOver, width = 0.33, height = 0.5, x = 0.67, y = 0.5, interpolate = TRUE )
```

\

Make sure to load the data and use inline code for some of this information.

## Questions

Question 1:

Do people tend to quantify acidity and bitterness in coffee correctly and is their ability to judge dependent on their preference in coffee beverage?

Question 2:

Does political affiliation play any part in coffee preference?

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

```{=html}
<table>
  <caption>
    <span data-qmd="Weekly Plan of Attack Table: Last updated: 5/28/2024">
    </span>
  </caption>
  
  <thead>
    <tr>
      <th>Task Name</th>
      <th>Status</th>
      <th>Assignee</th>
      <th>Due</th>
      <th>Priority</th>
      <th>Summary</th>
    </tr>
  </thead>
  <tr>
    <td>Add any extras</td>
    <td>in progress</td>
    <td>Cristina Lafuente</td>
    <td>Wednesday, May 29: 5pm</td>
    <td>Low</td>
    <td>Put in pictures, charts or plots to help viewers make sense of proposal</td>
  </tr>
  <tr>
    <td>Format and Beautify</td>
    <td>in progress</td>
    <td>Cristina Lafuente</td>
    <td>Wednesday, May 29: 5pm</td>
    <td>Moderate</td>
    <td>Go through and make sure that all formatting is nice </td>
  </tr>
  <tr>
    <td>Add req'd info to proposal</td>
    <td>in progress</td>
    <td>Cristina Lafuente</td>
    <td>Wednesday, May 29: 5pm</td>
    <td>High</td>
    <td>Fill in all required information </td>
  </tr>
</table>
```
