---
title: "The Great Coffee Caper"
subtitle: "Exploring Discernment and Appreciation of Different Roasts"
author: 
  - name: "Cristina Lafuente"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Evaluation of data recovered from coffee taste test"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: true
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
library(here)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

coffeeData <- readr::read_csv(('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-05-14/coffee_survey.csv'))
```

The Coffee Data comes from a coffee taste test held on YouTube in Oct 2023. Data was self reported by participants. It was aggregated by a data blogger named Robert McKeon Aloe who analyzed the data the following month.

The Dimensions of the data are:

```{r}
dim(coffeeData)
```

as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

## Questions

Question 1:

Do people tend to quantify acidity and bitterness in coffee as acidity and bitterness distinctly and correctly?

Question 2:

Does political affiliation play any part in coffee preference?

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

```{=html}
<table>
  <caption>
    <span data-qmd="Weekly Plan of Attack Table: Last updated: 5/21/2024">
    </span>
  </caption>
  
  <thead>
    <tr>
      <th>Task Name</th>
      <th>Status</th>
      <th>Assignee</th>
      <th>Due</th>
      <th>Priority</th>
      <th>Summary</th>
    </tr>
  </thead>
  <tr>
    <td>**</td>
    <td>**</td>
    <td>Cristina Lafuente</td>
    <td>Wednesday, May 29: 5pm</td>
    <td>Low</td>
    <td>**</td>
  </tr>
  <tr>
    <td>**</td>
    <td>**</td>
    <td>Cristina Lafuente</td>
    <td>Wednesday, May 29: 5pm</td>
    <td>Moderate</td>
    <td>**</td>
  </tr>
  <tr>
    <td>**</td>
    <td>**</td>
    <td>Cristina Lafuente</td>
    <td>Wednesday, May 29: 5pm</td>
    <td>High</td>
    <td>**</td>
  </tr>
</table>
```
