---
title: "The Great Coffee Caper"
subtitle: "Exploring Discernment and Appreciation of Different Roasts"
author: 
  - name: "Cristina Lafuente"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Evaluation of data recovered from coffee taste test"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  message: FALSE
  warning: FALSE
---

```{r}
#| label: load-pkgs
#| message: FALSE


if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(cowplot,
               dplyr,
               here,
               knitr,
               magick,
               scales,
               shiny,
               tidyverse,
               viridis)

devtools::install_github("tidyverse/dsbox")
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

coffeeData <- readr::read_csv(('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-05-14/coffee_survey.csv'))
```

The Coffee Data comes from a coffee taste test held on YouTube in Oct 2023. Data was self reported by participants. It was aggregated by a data blogger named Robert McKeon Aloe who analyzed the data the following month.

The Dimensions of the data are:

```{r}
dim(coffeeData)
```

The variables collected are:

```{r}
colnames(coffeeData)
```

The reason I selected this data set is because I enjoy coffee very much! Too much? Probably not. I have 6 different coffee brewing apparatuses in my kitchen.

```{r fig.cap = "Images from amazon.com reproduced here without permission"}

mokaPot <- (here("images/mokaPot.png"))
espressoMaker <- (here("images/espressoMachine.png"))
frenchPress <- (here("images/frenchPress.png"))
pod <- (here("images/kCup.png"))
percolator <- (here("images/percolator.png"))
pourOver <- (here("images/pourOver.png"))

ggdraw() +
  draw_image(mokaPot, width = 0.33, height = .5, y = 0, interpolate = TRUE) +
  draw_image(espressoMaker, width = 0.33, height = 0.5, x = 0.33, y = 0, interpolate = TRUE) +
  draw_image(frenchPress, width = 0.33, height = 0.5, x = 0.67, y = 0, interpolate = TRUE) + 
  draw_image(pod, width = 0.33, height = 0.5, y = 0.5 , interpolate = TRUE) +
  draw_image(percolator, width = 0.33, height = 0.5, x = 0.33, y = 0.5, interpolate = TRUE) + 
  draw_image(pourOver, width = 0.33, height = 0.5, x = 0.67, y = 0.5, interpolate = TRUE )
```

Looking at data on coffee for six weeks seemed like a natural choice.

## Questions

Question 1:

Do people tend to quantify acidity and bitterness in coffee correctly and is their ability to judge dependent on their preference in coffee beverage?

Question 2:

Does political affiliation play any part in coffee preference?

## Analysis plan

-   Additional data will come from: [Acids in coffee: A review of sensory measurements and meta-analysis of chemical composition, by Yeager et al](https://www.tandfonline.com/doi/figure/10.1080/10408398.2021.1957767?)\

-   There are two relevant CSVs one contains data which will be used to describe contr

    This data will be joined on the type of roast (variable in both).

    This study provides definitive data on the levels of acidity in coffee by roast as well as a source for what causes both bitterness and acidity in taste

    ```{r, error = FALSE, message = FALSE, echo = FALSE}
    coffeeAcidityCGA <- readr::read_csv(here("data/Acids_in_Coffee_-CGAs.csv"))
    coffeeAcidityOA <- readr::read_csv(here("data/Acids_in_Coffee_-OAs.csv"))
    ```

They have quite a few very nice plots of their own, I think the one that best summarizes the data I'll be using is:

```{r fig.cap = "Image from Yeager et al, Acids in Coffee"}
plotAcidByRoast <- here("images/acidLevelByRoast.png")

ggdraw() +
  draw_image(plotAcidByRoast)

```

Due to some type of formatting error, roughly 1500 and 250 NA columns were appended to each dataset. Trimming those off, allows a more realistic view of the actual data names and size.

This new data for chlorogenic acids has the dimensions and columns:

```{r}
## Dimensions and columns in the chlorogenic acids dataset:
dim(coffeeAcidityCGA[1:33])

```

The columns are called:

```{r}
## The columns are called:
colnames(coffeeAcidityCGA[1:33])

```

The new data for organic acids has dimension:

```{r}

## The dimensions of the Organic Acids dataset is:
dim(coffeeAcidityOA[1:33])
```

That data has column names:

```{r}
## Those columns are called:
colnames(coffeeAcidityOA[1:33])
```

Working with this additional data will require that I combine these three datasets along the "roast" variable and compare the relevant variables.

For Organic Acids, I will need to create a new variable which adds up the total relevant organic acids.

In the CGAs Table, I will similarly need to create a new variable which sums the total relevant CGAs.

Once this has been done, those two variables can be compared, by roast, against what the taste testers believed to be acidic and non acidic.

After that, I would like to run an analysis on the likelihood of a person to be correct based on their favorite coffee beverage.

Looking at the second question, will be much more straightforward examining the political affiliation selection made by the participants as well as which of the coffees they preferred.

```{=html}
<table border ="1">
  <caption>
    <span data-qmd="Weekly Plan of Attack Table: Last updated: 5/28/2024">
    </span>
  </caption>
  
  <thead style =" background-color: #BF93EE">
    <tr>
      <th>Task Name</th>
      <th>Status</th>
      <th>Assignee</th>
      <th>Due</th>
      <th>Priority</th>
      <th>Summary</th>
    </tr>
  </thead>
  <tr>
    <td>Add any extras or polish up wording</td>
    <td>not yet started</td>
    <td>Cristina Lafuente</td>
    <td>June 10, 2024</td>
    <td>Low</td>
    <td>As possible additions *nice to haves* pop up, add them to this list</td>
  </tr>
  <tr>
    <td>Begin working on presentation and write up</td>
    <td>not yet started</td>
    <td>Cristina Lafuente</td>
    <td>June 3, 2024</td>
    <td>Moderate</td>
    <td>Once feedback is in, begin wrangling data and working on presentation</td>
  </tr>
  <tr>
    <td>Finish proposal and give feedback</td>
    <td>nearly complete</td>
    <td>Cristina Lafuente</td>
    <td>Wednesday, May 29: 5pm</td>
    <td>High</td>
    <td>Push completed proposal and complete peer review </td>
  </tr>
</table>
```
